Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens

### 2014-07-25 limits/ms.tex
Justin's changes complete. New referee reply file mostly complete. Action items (very minor) that remain marked in ./mnras/rev1-comments_reply.txt with 1120$

%\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\title[Lens Recovery with \Glass]{Stellar Mass and Stellar Kinematics Can Constrain Gravitational Lens Models}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$45 avenue des Etats-Unis, 78035, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

\keywords{gravitational lensing: strong, methods: numerical, methods: statistical}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing (see also \S\ref{sec:theory} for a more detailed exposition). First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology\footnote{Here, $c$ is the speed of light in vacuo and $H_0$ is the Hubble parameter.}.  Typically:
%
\begin{equation}
D_L \approx z_L \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
where $z_L$ is the redshift of the lens. 
%
For multiple images, the sky-projected density must exceed the critical
density in some region:
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
where $G$ is Newton's gravitational constant. 
%
The angular separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$)
is the gravitational radius. If the source is a quasar or otherwise rapidly
variable, a time delay $\Delta t$ in the variability will be present where:
%
\begin{equation}
\Delta t \sim R_G/c
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\item {\bf Model ensembles:} Model ensembles, exploring a diverse range of
    possible mass distributions that nonetheless all fit the data, are a way of
    combating the non-uniqueness of models.  Such ensembles are possible in
    parametric models \citep{1999AJ....118...14B} but are more common in
    free-form models, where -- since such models are deliberately
    under-constrained -- they become vital
    \citep{2000AJ....119..439W,2009ApJ...690..154S,2012MNRAS.425.3077L}.

@@

In this paper we demonstrate which combination of lensing, stellar mass, and
stellar kinematic constraints best constrain the projected mass profile and
shape of several gravitational lens produced from mock data. In particular,
we introduce the idea of using estimates of the stellar mas content to place
lower bounds on the lensing mass.
We have also developed a new lens modeling framework  -- \Glass\ (Gravitational
Lensing AnalysiS Software) which improves upon the earlier techniques. The
key developments are

\begin{enumerate}
\item A modular framework that allows new priors to be added and modified easily.
\item The basis functions approximating a model can be changed.
\item Non-linear constraints can be imposed in an automated post-processing step.
\item The central region can have a higher resolution to capture steep models.
\item Stellar density can be used as an additional constraint.
\item Point or extended mass objects can be placed in the field.
\item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}

The compute-intensive core of \Glass\ is a parallel
implementation of the high-dimensional sampling algorithm from
\cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has extensible code (in
software terminology, a framework) in Python, encoding the lens equation, a
basis and priors for the mass distribution, and various post-processing and
visualisation modules.

\Glass\ is described in \secref{sec:glass}.  In \secref{sec:theory}, we review
the key elements of lensing theory, stellar population synthesis, and stellar
dynamics we will need. In \secref{sec:mockdata}, we describe our mock data. In
\secref{sec:results}, we present our results from applying \Glass\ to these
mock data. Finally, in \secref{sec:conclusions} we present our conclusions.

@@

The rather complicated dependence of lensing observables on the mass
distribution $\kappa(\vec\theta)$ has an important consequence: very different
mass distributions can result in similar observables.  This is the phenomenon
of lensing degeneracies.  While the non-uniqueness of lens models noted by
\cite{1981ApJ...244..736Y} already hinted at degeneracies, their existence was
first derived by \cite{1985ApJ...289L...1F}.  The most important is the
so-called mass-sheet degeneracy, which is that image positions remain invariant
if $\tau(\vec\theta)$ is multiplied by an arbitrary constant.  
\hilight{This corresponds
to rescaling the surface density at the images $\kappa(\vec\theta)$.} In fact
there are infinitely many degeneracies \citep{2000AJ....120.1654S} because any
transformation of the arrival-time surface away from the images has no effect
on the lensing observables.  In particular, there are degeneracies that involve
the shape of the mass distribution
\citep{2006ApJ...653..936S,2013arXiv1306.4675S}.  Degeneracies tend to be
suppressed if there are sources at very different redshifts or `redshift
contrast' \citep{1998AJ....116.1541A,2009ApJ...690..154S}, because the presence
of different factors of $D_S/D_{LS}$ in the image plane makes it more difficult
to change the mass distribution and the arrival-time surface without affecting
the lensing observables.  But degeneracies are still present with multiple
source redshifts.  \citep{2008MNRAS.386..307L}.

@@

The key scientific and technical improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework that allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on
either a single lens object or the entire ensemble of objects. \Glass\ comes
with a number of useful priors, the default ones will be described later, but a
user can write their own directly in the input file, or by modifying the source
code.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the
code has been designed to support alternative methods. In particular, there are
future plans to develop a module using Bessel functions. This will require a
new set of priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints,
any number of post processing functions can be applied. Not only can these
functions be used to derive new quantities from the mass models, they can also
be used as a filter to accept or reject a model based on some non-linear
constraint. The plotting functions within \Glass\ will correctly display models
that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can have a higher resolution to capture steep models. 
\end{enumerate}
%
With the default basis set of pixels, the mass distribution of the lens is
described by a uniform grid. However, in the central region of a lensing galaxy
where the mass profile may rise steeply, the center pixel uses a higher
resolution. This allows the density to increase smoothly but still allow for a
large degree of freedom within the inner region without allowing the density to
be arbitrarily high. 
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
A shear term can be added to the potential, as shown later in \eqnref{shear},
to account for mass external to the modelled region. This is useful to capture
the gross effects of a distant neighbour, since there is a degeneracy between
the ellipticity of a lens and shear field: The greater the allowed shear, the
more circular the lens may be.  \Glass\ also allows further analytic potential
components to be included. This can be used to model substructure or multiple
neighbours close to the main lens. The substructure may have only a small
effect if the lens is a single galaxy, but if the lens is a group then a
potential can be added for each of the galaxies. A few standard functions are
already included in \Glass\ including those for a point mass, a power law
distribution, or an isothermal (a particular case of the power law).
%
\begin{enumerate}
  \setcounter{enumi}{6}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{2012MNRAS.425.3077L}; it is multi-threaded
allowing it to run efficiently on many-cored machines.

@@

\subsection{Discrete models}\label{sec:discrete}
For this paper, we will restrict ourselves to using a pixelated basis set that
has been employed in previous work. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points are solutions to the lens equation and satisfy other physically
motivated linear priors.  We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
\eqnrefp{lensing potential} as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the
logarithm over pixel $n$. The exact form for $Q$ is described in \appref{Q
derivation}.  We can find the discretized lens equation by simply taking the
gradient of the above equations.

The pixels only cover a finite circular area with physical radius $\Rmap$ and
pixel radius $\Rpix$ with the central cell centered on the lensing galaxy. To
account for any global shearing outside this region from, e.g., a neighboring
galaxy, we also add to \eqnref{discrete potential} two shearing terms
%
\begin{equation}
\label{shear}
\gamma_1(\theta_x^2 - \theta_y^2) + 2\gamma_2\theta_x\theta_y\quad.
\end{equation}
%
We can continue adding terms to account for other potentials. For instance, we
may want to impose a base potential over the field, or add potentials from the
presence of other galaxies in the field. \Glass\ already includes potentials
for a point mass or an exponential form, but custom potentials are
straightforward to add and can be included directly in the input file.  If the
stellar density $\kappa_s$ has been estimated we can use this as a lower bound
where the stellar potential is a known constant of the form \eqnref{discrete
potential}, e.g., $\kappa_n = \kappa_{\mathrm{dm},n} + \kappa_{s,n}$ for a
two-component model.

The lens equation and the arrival times alone are not enough to form a closed
volume in the solution space. The data alone do not even constraint the density
to be non-negative everywhere. We therefore provide additional linear
constraints so that only physical models are sampled. 
\Glass\ provides a number of typical constraints similar to those discussed in
\cite{2008ApJ...679...17C}: 
%
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The azimuthally averaged density profile must have a slope everywhere $\le 0$.
\item Image parity is enforced.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item The density is inversion symmetric.
\end{enumerate}

The first three of these priors are very important for creating physically
meaningful lenses. Prior (ii) is necessary to prevent rings from forming where
the inner region is less dense. This would be unusual for a galaxy or a
cluster. Prior (iv) can be relaxed somewhat by increasing the angle. The
smoothness prior (v) requires pixel values to be less than a constant factor
(normally 2) of the average neighbor value. This can be relaxed by increasing
the constant or completely disabled. For this paper, we have chosen \emph{not}
to use this prior, and therefore allow for the presence of small density
substructure.  Prior (vi) enforces a radial symmetry that requires radially
opposed pixels (i.e., on opposite sides of the center) to be equal. This prior
is particularly important for modelling doubles which must be symmetric. If a
lens is suspected to be symmetric this can also significantly improve the
modelling time for quads.

While most of these priors have been selected for modelling galaxies, there
are several ways to address clusters or allow for the presence of substructure.
As previously described, one can add any number of additional analytic 
potentials inside or outside the main modelling pixelation area. For instance,
in a cluster, one can explicitly account for galaxies with power law potentials.
Furthermore, Priors (iv) and (v) and be relaxed or disabled. We have perfomed
preliminary testing of these ideas several common clusters and are encouraged
by the results. We will reserve a full analysis for a future publication.

In the simplest form, a single model for a lens is a tuple $\M = (\vec\kappa,
\vec\beta, \gamma_1, \gamma_2)$. A single model represents a single point in
the solution space polytope. Using the MCMC sampling strategy described in
\cite{2012MNRAS.425.3077L} we uniformly sample this space. Collectively, the
sampled models are referred to as an ensemble $\E = \{\M_i\}$, where we usually
generate $|\E| \sim 1000$ models. One can choose to further process these
models to impose priors that may be difficult to enforce during the
modeling process. For instance, non-linear constraints, or simply filtering
of models that do not meet some criteria can be excluded, or weighted against.
For the current analysis, we do not exclude any models and treat all models
as equally likely.

The time to generate the model ensemble is mostly a function of the size of the
parameter space. The MCMC algorithm has a ``warm-up'' phase where it estimates
the size and shape of each dimension in the solution space.  Once this has been
completed, the models are sampled very quickly. In fact, there is little
difference between generating 1,000 or 10,000 models, although we find little
statistical difference after 1,000 models. For the mock lenses, the typical
``warm-up'' time was about 4s, and the modelling time was 20s using a parallel
shared-memory machine with 40 cores.

@@

\begin{figure*}
\includegraphics[width=0.33\textwidth]{MockGalProfile-a.pdf} 
\includegraphics[width=0.33\textwidth]{MockGalProfile-b.pdf} 
\includegraphics[width=0.33\textwidth]{MockGalProfile-c.pdf}
\caption{
Profiles of the four mock galaxies showing the stellar (dotted) and dark matter (dashed) components and the total (solid).
\textbf{Left:} 
The spherically averaged density. The stars in models \mockBB{} and \mockBC{} contribute significantly to the central potential. 
\textbf{Middle:} 
The radially averaged two-dimensional projected density.
The critical density for lensing at $z_L=0.31$, $\kappa_\mathrm{crit}\sim 1.8\e{9}$\Msun/kpc$^2$, is marked by the horizontal line. 
\textbf{Right:}
The enclosed projected mass.
}
\label{mock galaxies}
\end{figure*}

@@

\begin{enumerate}
\item one double and one extended double;
\item one quad and one extended quad;
\item two 2-source quads with varying redshift contrast.
\end{enumerate}
The `extended' configurations use multiple point sources at the same redshift
to simulate an extended source that will produces an arc-like image.
\figref{arrival surfaces} shows the lens configurations for the \mockBC\
galaxy. 
The configurations for the other galaxies are similar.  The labels Z1, Z2, Z3
within the names refer to the redshift of the sources. We have chosen Z1=1.72,
Z2=0.72, and Z3=0.51 so that the radial distribution of the images is roughly
equally spaced. For all mocks, we do not apply any external shear field.  Only
the central image of the Z1 source is used to avoid overconstraining the
models, otherwise all the central images would fall within the central pixel
and no solution exists that satisfies all locations simultaneously for one
pixel value.

\begin{figure*}
\includegraphics[width=0.85\textwidth]{BCarrival_surfaces}
\caption{The lens configurations for the six test cases using the \mockBC{}
mock galaxy. The other mock galaxies produce similar results. Here, the central
image is shown, although not all tests include it. The naming convention
indicates the redshift of the sources with Z1=1.72, Z2=0.72, and Z3=0.51.  The
central image only belongs to the Z1 source to avoid overconstraining the
models. Images of the same color share a common source. The extended source examples have been
contructed so that the images will form arclets.
Grey circles are a visual aid
to help determine radial separation between images. The axes are in
arcseconds.}
\label{arrival surfaces}
\end{figure*}

>>>>

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
%\title[Lens Recovery with \Glass]{Stellar Mass and Stellar Kinematics Can Constrain Gravitational Lens Models}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$45 avenue des Etats-Unis, 78035, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

\begin{keywords}
gravitational lensing: strong, methods: numerical, methods: statistical
\end{keywords}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing (see also \S\ref{sec:theory} for a more detailed exposition). First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology\footnote{Here, $c$ is the speed of light in vacuo and $H_0$ is the Hubble parameter.}.  Typically:
%
\begin{equation}
D_L \approx z_L \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
where $z_L$ is the redshift of the lens. 
%
For multiple images, the sky-projected density must exceed the critical lensing density in some region:
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
where $G$ is Newton's gravitational constant. 
%
The angular separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$)
is the gravitational radius. If the source is a quasar or otherwise rapidly
variable, a time delay $\Delta t$ in the variability will be present where:
%
\begin{equation}
\Delta t \sim R_G/c
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\item {\bf Model ensembles:} Model ensembles, exploring a diverse range of
    possible mass distributions that nonetheless all fit the data, are a way of
    combating the non-uniqueness of models.  Such ensembles are possible in
    parametric models \citep[e.g.][]{1999AJ....118...14B,2010Sci...329..924J} but are more common in
    free-form models, where -- since such models are deliberately
    under-constrained -- they become vital
    \citep{2000AJ....119..439W,2009ApJ...690..154S,2012MNRAS.425.3077L}.

@@

In this paper, we introduce a new non-parametric lens modelling framework  -- \Glass\ (Gravitational
Lensing AnalysiS Software). This shares some aspects with an earlier code \PixeLens \citep{Saha2004,2008ApJ...679...17C}. However, \Glass\ -- which contains all new code written from the ground up -- significantly improves upon \PixeLens\ in several key ways:

\begin{enumerate}
\item At the heart of \Glass\ is a new uniform sampling algorithm for high dimensional spaces \citep{2012MNRAS.425.3077L}. This allows for large ensembles of $>10,000$ models to be efficiently generated. 
\item \Glass\ provides a modular framework that allows new priors to be added and modified easily.
\item The basis functions approximating a model can be easily changed (in this paper, we assume pixels as in \PixeLens). 
\item With so many models in the final ensemble, we can afford to apply non-linear constraints (for example stellar kinematic data; or the removal of models with spurious extra images) to accept/reject models in a post-processing step.
\item The central region of the mass map can have a higher resolution to more efficiently capture steep models.
\item Stellar density can be used as an additional constraint on the models. 
\item Point or extended mass objects can be placed in the field.
\end{enumerate}
As a first application, we use \Glass\ on mock data to determine which combination of lensing, stellar mass and/or 
stellar kinematic constraints best constrain the projected mass profile and
shape of a gravitational lens. We will apply \Glass\ to real lens data in a series of forthcoming papers.

%The compute-intensive core of \Glass\ is a parallel
%implementation of the high-dimensional sampling algorithm from
%\cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has extensible code (in
%software terminology, a framework) in Python, encoding the lens equation, a
%basis and priors for the mass distribution, and various post-processing and
%visualisation modules.

This paper is organised as follows. In \secref{sec:glass}, we describe the \Glass\ code. In \secref{sec:theory}, we review
the key elements of lensing theory, stellar population synthesis, and stellar
dynamics we will need. In \secref{sec:mockdata}, we describe our mock data. In
\secref{sec:results}, we present our results from applying \Glass\ to these
mock data. Finally, in \secref{sec:conclusions} we present our conclusions.

@@

The rather complicated dependence of lensing observables on the mass
distribution $\kappa(\vec\theta)$ has an important consequence: very different
mass distributions can result in similar observables.  This is the phenomenon
of lensing degeneracies.  While the non-uniqueness of lens models noted by
\cite{1981ApJ...244..736Y} already hinted at degeneracies, their existence was
first derived by \cite{1985ApJ...289L...1F}.  The most important is the
so-called mass-sheet degeneracy, which is that image positions remain invariant
if $\tau(\vec\theta)$ is multiplied by an arbitrary constant.  
\hilight{This corresponds
to rescaling the surface density at the images $\kappa(\vec\theta)$.} In fact
there are infinitely many degeneracies \citep{2000AJ....120.1654S} because any
transformation of the arrival-time surface away from the images has no effect
on the lensing observables.  In particular, there are degeneracies that involve
the shape of the mass distribution
\citep{2006ApJ...653..936S,2013arXiv1306.4675S}.  Degeneracies tend to be
suppressed if there are sources at very different redshifts or `redshift
contrast' \citep{1998AJ....116.1541A,2009ApJ...690..154S}, because the presence
of different factors of $D_S/D_{LS}$ in the image plane makes it more difficult
to change the mass distribution and the arrival-time surface without affecting
the lensing observables. But degeneracies are still present with multiple
source redshifts \citep{2008MNRAS.386..307L,2014arXiv1406.6152S}.

@@

The key scientific and technical improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A new uniform sampling algorithm for high dimensional spaces.
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{2012MNRAS.425.3077L}; it is multi-threaded
allowing it to run efficiently on many-cored machines.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item A modular framework that allows new priors to be added and modified easily.
%
Each prior is a simple function that adds linear constraints that operate on
either a single lens object or the entire ensemble of objects. \Glass\ comes
with a number of useful priors (the default ones will be described in \secref{sec:discrete}), but a
user can write their own directly in the input file, or by modifying the source
code.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item The basis functions approximating a model can be changed. 
%
\Glass\ currently describes the lens mass as a collection of pixels, but the
code has been designed to support alternative methods. In particular, there are
future plans to develop a module using Bessel functions. This will require a
new set of priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
%
Once \Glass\ has generated an ensemble of models given the linear constraints,
any number of post processing functions can be applied. Not only can these
functions be used to derive new quantities from the mass models, they can also
be used as a filter to accept or reject a model based on some non-linear
constraint. For example, we can reject models that have spurious extra images (\secref{sec:glassextraimages}), or models that do not match stellar kinematic constraints (\secref{sec:glasskinematics}). The plotting functions within \Glass\ will correctly display models
that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item The central region can have a higher resolution to capture steep models. 
%
With the default basis set of pixels, the mass distribution of the lens is
described by a uniform grid. However, in the central region of a lensing galaxy
where the mass profile may rise steeply, the center pixel uses a higher
resolution. This allows the density to increase smoothly but still allow for a
large degree of freedom within the inner region without allowing the density to
be arbitrarily high. 
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Stellar density can be used as an additional constraint.  
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{6}
  \item Point or extended mass objects can be placed in the field.
%
A shear term can be added to the potential, as shown later in \eqnref{shear},
to account for mass external to the modelled region. This is useful to capture
the gross effects of a distant neighbour, since there is a degeneracy between
the ellipticity of a lens and its shear field (the greater the allowed shear, the
more circular the lens may be). \Glass\ also allows further analytic potential
components to be included. These can be used to model substructure or multiple
neighbours close to the main lens. The substructure may have only a small
effect if the lens is a single galaxy, but if the lens is a group or cluster then a
potential can be added for each of the known member galaxies. A few standard functions are
already included in \Glass\ including those for a point mass, a power law
distribution, or an isothermal (a particular case of the power law).

@@

\subsection{Pixelated models}\label{sec:discrete}
For this paper, we will restrict ourselves to using a pixelated basis set as used by \PixeLens\ \citep{Saha2004,2008ApJ...679...17C}, but note that it is straightforward to add other basis function expansions to \Glass. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points satisfy both the lens equation and other physically
motivated {\it linear} priors \citep{2012MNRAS.425.3077L}. A limitation of our sampling strategy is that only linear constraints may be applied when building the model ensemble; however, non-linear constraints can be applied in post-processing (see \secref{sec:glassextraimages} and \secref{sec:glasskinematics}). We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
\eqnrefp{lensing potential} as:
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the
logarithm over pixel $n$. The exact form for $Q$ is described in \appref{Q
derivation}.  We can find the discretized lens equation by simply taking the
gradient of the above equations.

The pixels only cover a finite circular area with physical radius $\Rmap$ and
pixel radius $\Rpix$ with the central cell centered on the lensing galaxy. To
account for any global shearing outside this region from, e.g., a neighboring
galaxy, we also add to \eqnref{discrete potential} two shearing terms:
%
\begin{equation}
\label{shear}
\gamma_1(\theta_x^2 - \theta_y^2) + 2\gamma_2\theta_x\theta_y\quad.
\end{equation}
%
We can continue adding terms to account for other potentials. For instance, we
may want to impose a base potential over the field, or add potentials from the
presence of other galaxies in the field. \Glass\ already includes potentials
for a point mass or an exponential form, but custom potentials are
straightforward to add and can be included directly in the input file.  If the
stellar density $\kappa_s$ has been estimated we can use this as a lower bound
where the stellar potential is a known constant of the form \eqnref{discrete
potential}, e.g., $\kappa_n = \kappa_{\mathrm{dm},n} + \kappa_{s,n}$ for a
two-component model.

\subsubsection{Priors}

The lens equation and the arrival times alone are typically not enough to form a closed
volume in the solution space. We therefore require additional linear
constraints -- {\it priors}. Some of these are `physical' in the sense that they are unarguable -- for example demanding that the mass density is everywhere positive; others are more subjective, for example demanding that the mass map is smooth over some region. Such `regularisation' priors may be switched off for all or some of the mass map if the data are sufficiently constraining.

The priors built in to \Glass\ are similar to those used in \PixeLens\ \citep{2008ApJ...679...17C}. The physical priors are always used by default; the regularisation priors are used sparingly -- i.e. only if the data are not sufficiently constraining to obtain sensible solutions without them:

%
\vspace{2mm}
\noindent
{\bf Physical priors}
\begin{enumerate}
\item The density must be non-negative everywhere.
\item Image parity is enforced.
\end{enumerate}

\vspace{1mm}
\noindent
{\bf Regularisation priors}
\begin{enumerate}
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item The azimuthally averaged density profile must have a slope everywhere $\le 0$.
\item The density is inversion symmetric.
\end{enumerate}
For typical lens data, the regularisation priors are very important for creating physically
sensible solutions. Prior (i) demands that the peak in the mass density is at the centre of the mass map. Secondary `plateaus' in the mass map are possible, but not secondary peaks. Note that this prior still successfully allows merging galaxy systems to be correctly captured, provided that the two galaxies are not equally dense in projection (see, for example the \PixeLens\ model of the merger system B1608 in \citealt{2007ApJ...667..645R}); and for the successful detection of `meso-structure' in strong lensing galaxy clusters \citep{2007ApJ...663...29S}. Prior (ii) is arguably a physical prior sense negative azimuthally averaged density slope would be unstable \citep[e.g.][]{2008gady.book.....B}. Note that this prior does not preclude successful modelling of mergers or substructure unless the total projected mass in substructure is comparable to the projected mass of the host in an azimuthal annulus \citep{2007ApJ...667..645R,2007ApJ...663...29S}. Prior (iii) is only used for doubles that ought to be inversion symmetric and quads where inversion symmetry is clear from the image configuration.

Finally, we remind the reader that all of the regularisation priors can be switched off or changed/improved depending on the data quality available. For clusters, substructure can be explicitly modelled by adding analytic potentials at the known locations of galaxies; furthermore the above priors can be relaxed in regions of the mass map where the data are particularly constraining (for example near the images). We will apply \Glass\ to a host of strong lensing clusters in forthcoming work, where we will explicitly test the prior on mock data that has significant substructure.

\subsection{Building the model ensemble}

In the simplest form, a single model for a lens is a tuple $\M = (\vec\kappa,
\vec\beta, \gamma_1, \gamma_2)$. A single model represents a single point in
the solution space polytope. Using the MCMC sampling strategy described in
\cite{2012MNRAS.425.3077L} we uniformly sample this space. Collectively, the
sampled models are referred to as an ensemble $\E = \{\M_i\}$, where we usually
generate $|\E| \sim 1000$ models. One can choose to further process these
models to impose priors that may be difficult to enforce during the
modeling process. For instance, non-linear constraints, or simply filtering
of models that do not meet some criteria can be excluded, or weighted against as discussed previously.
In this paper, we do not exclude any models and treat all models
as equally likely.

The time to generate the model ensemble is mostly a function of the size of the
parameter space. The MCMC algorithm has a ``warm-up'' phase where it estimates
the size and shape of each dimension in the solution space.  Once this has been
completed, the models are sampled very quickly. In fact, there is little
difference between generating 1,000 or 10,000 models, although we find little
statistical difference after 1,000 models. For the mock lenses, the typical
``warm-up'' time was about 4s, and the modelling time was 20s using a parallel
shared-memory machine with 40 cores. The ability to rapidly generate so many models is what allows us to then accept/reject models to apply non-linear constraints (see \secref{sec:glassextraimages} and \secref{sec:glasskinematics}). This is a key advantage over our earlier pixelated strong lens tool \PixeLens.

@@

\begin{figure*}
\includegraphics[width=0.33\textwidth]{MockGalProfile-a.pdf} 
\includegraphics[width=0.33\textwidth]{MockGalProfile-b.pdf} 
\includegraphics[width=0.33\textwidth]{MockGalProfile-c.pdf}
\caption{
Profiles of the four mock galaxies showing the stellar (dotted) and dark matter (dashed) components and the total (solid).
\textbf{Left:} 
The spherically averaged density. The stars in models \mockBB{} and \mockBC{} contribute significantly to the central potential. 
\textbf{Middle:} 
The radially averaged two-dimensional projected density.
The critical lensing density at $z_L=0.31$, $\kappa_\mathrm{crit}\sim 1.8\e{9}$\Msun/kpc$^2$, is marked by the horizontal line. 
\textbf{Right:}
The enclosed projected mass.
}
\label{mock galaxies}
\end{figure*}

@@

\begin{enumerate}
\item one double and one extended double;
\item one quad and one extended quad;
\item two 2-source quads with varying redshift contrast.
\end{enumerate}
The `extended' configurations use multiple point sources at the same redshift
to simulate an extended source that will produces an arc-like image.
\figref{arrival surfaces} shows the lens configurations for the \mockBC\
galaxy. 
The configurations for the other galaxies are similar.  The labels Z1, Z2, Z3
within the names refer to the redshift of the sources. We have chosen Z1=1.72,
Z2=0.72, and Z3=0.51 so that the radial distribution of the images is roughly
equally spaced. For all mocks, we do not apply any external shear field.  Only
the central image of the Z1 source is used to avoid over-constraining the
models, otherwise all the central images would fall within the central pixel
and no solution exists that satisfies all locations simultaneously for one
pixel value.

\begin{figure*}
\includegraphics[width=0.85\textwidth]{BCarrival_surfaces}
\caption{The lens configurations for the six test cases using the \mockBC{}
mock galaxy. The other mock galaxies produce similar results. Here, the central
image is shown, although not all tests include it. The naming convention
indicates the redshift of the sources with Z1=1.72, Z2=0.72, and Z3=0.51.  The
central image only belongs to the Z1 source to avoid over-constraining the
models (see \secref{sec:lensconfig} for further details). Images of the same colour share a common source. The extended source examples have been constructed so that the images will form arclets.
Grey circles are a visual aid
to help determine radial separation between images. The axes are in
arcseconds.}
\label{arrival surfaces}
\end{figure*}
